\documentclass[../RFC-HDFG-2026-001.tex]{subfiles}

\begin{document}

%% ======================================================================
%%  Section 7 – Built-in Filter Parameters
%% ======================================================================
\section{Built-in Filter Parameters}
\label{sec:builtin-filters}

Each built-in filter implements \texttt{set\_config} and \texttt{get\_config} so that the
string API is fully functional without third-party plugins.

For built-in filters that have dedicated \texttt{H5Pset\_*} functions with non-trivial
validation logic (particularly SZIP and Scale-Offset), the \texttt{set\_config}
callback delegates to the existing internal validation code rather than
reimplementing it independently. This prevents divergence between the two
configuration paths and ensures that validation constraints are applied
consistently regardless of whether the filter was configured via
\texttt{H5Pset\_szip} or \texttt{H5Pset\_filter\_params}.

\verysubsection{Implementation note}
Built-in filters are currently registered internally as \texttt{H5Z\_class2\_t}
structs. Each built-in filter's internal registration is converted to
\texttt{H5Z\_class3\_t} with \texttt{set\_config} and \texttt{get\_config} added directly to the
struct definition (e.g., in \texttt{H5Zdeflate.c}, \texttt{H5Zszip.c}, etc.).

This conversion is entirely internal to the library. External code never
directly accesses these struct definitions, so there is no public API
compatibility concern. The converted structs are passed to the existing
\texttt{H5Zregister()} path, which already handles \texttt{H5Z\_class3\_t} registrations
(\SectionRef{sec:class3}). No parallel mechanism or additional internal API is
required.

\subsection{Deflate}

\begin{itemize}
\item \texttt{level} (integer 0--9, default 6): Compression level.
\item Example: \texttt{"level=6"}
\end{itemize}

\subsection{Shuffle}

No parameters. \texttt{params} should be \texttt{NULL}.

\subsection{Fletcher32}

No parameters. \texttt{params} should be \texttt{NULL}.

\subsection{SZIP}

\begin{itemize}
\item \texttt{coding} (\texttt{"entropy"} or \texttt{"nn"}, default \texttt{"nn"}): Coding
  method.
\item \texttt{pixels\_per\_block} (even integer 2--32, default 16): Pixels per block.
\item The \texttt{set\_config} callback translates these into the appropriate option
  masks (\texttt{H5\_SZIP\_EC\_OPTION\_MASK} or \texttt{H5\_SZIP\_NN\_OPTION\_MASK}) and
  delegates to the existing \texttt{H5Pset\_szip} validation path. The existing
  \texttt{H5Pset\_szip} (\texttt{H5Pdcpl.c}) performs the following operations beyond
  simple parameter translation: checks whether the SZIP encoder is available
  (\texttt{H5Z\_FILTER\_CONFIG\_ENCODE\_ENABLED}); validates \texttt{pixels\_per\_block} is
  even and does not exceed \texttt{H5\_SZIP\_MAX\_PIXELS\_PER\_BLOCK}; applies
  \texttt{H5\_SZIP\_ALLOW\_K13\_OPTION\_MASK}; sets \texttt{H5\_SZIP\_RAW\_OPTION\_MASK};
  strips \texttt{H5\_SZIP\_CHIP\_OPTION\_MASK}, \texttt{H5\_SZIP\_LSB\_OPTION\_MASK}, and
  \texttt{H5\_SZIP\_MSB\_OPTION\_MASK}; and forces \texttt{H5Z\_FLAG\_OPTIONAL} regardless
  of the caller's flag choice. The \texttt{set\_config} callback replicates all
  of this behavior. Consequently, the \texttt{flags} argument to
  \texttt{H5Pset\_filter\_params} is overridden for SZIP---the filter is always
  optional. This is existing behavior of \texttt{H5Pset\_szip} and should be
  documented but is not changed by this feature. Callers that need to
  detect the flag override can compare the flags they passed with the
  flags returned by \texttt{H5Pget\_filter2} after the call.
\item Example: \texttt{"coding=entropy, pixels\_per\_block=8"}
\end{itemize}

\subsection{N-Bit}

No user-facing parameters (derived from datatype). \texttt{params} should be
\texttt{NULL}.

\subsection{Scale-Offset}

\begin{itemize}
\item \texttt{scale\_type} (\texttt{"float\_dscale"}, \texttt{"float\_escale"}, or
  \texttt{"int"}): Scale type. Required.
\item \texttt{scale\_factor} (integer): Scale factor. Required.
\item The \texttt{set\_config} callback delegates to the existing
  \texttt{H5Pset\_scaleoffset} validation path.
\item Example: \texttt{"scale\_type=float\_dscale, scale\_factor=4"}
\end{itemize}


%% ======================================================================
%%  Section 8 – Thread Safety and Concurrency
%% ======================================================================
\section{Thread Safety and Concurrency}
\label{sec:thread-safety}

\subsection{Plugin Loading Protection}

The on-demand plugin scan triggered by \texttt{H5PL\_load} performs filesystem
I/O (directory iteration and shared-library opens). In thread-safe builds of
HDF5 (configured with \texttt{--enable-threadsafe}), this I/O is serialized
under the global API lock.

The existing HDF5 global API lock serializes all public API entry points.
This is sufficient to prevent concurrent plugin loading races from user-facing
calls. No new locking mechanisms are introduced.

\verysubsection{Future-proofing note}
When many threads simultaneously encounter an unknown filter ID at startup,
all threads queue behind the first thread performing the plugin scan, producing
a convoy effect. Reducing lock granularity for the scan itself---for example
by allowing concurrent read-only scans while only serializing the final plugin
table insertion---would benefit throughput if the library moves toward
finer-grained concurrency.

\subsection{Parallel HDF5 (MPI) Considerations}

In Parallel HDF5, all MPI ranks participating in collective dataset creation
present identical dataset creation property lists (DCPLs). Non-deterministic
plugin loading across ranks (e.g., due to filesystem latency or differing
\texttt{HDF5\_PLUGIN\_PATH} settings) could cause one rank to fail to load a
required plugin while others succeed.

This is not a new problem introduced by this design---the same class of
issue exists today when any rank fails to load a required plugin, and there
are already several runtime-specific mechanisms (environment variables,
plugin path configuration, dynamic library search order) that can cause
per-rank divergence.

An additional parallel-specific concern is that parameter string parsing is
performed independently on every rank, even though the resulting
\texttt{cd\_values} are identical across ranks and the work need only be done
once.  This is an inherent inefficiency of the current design, though it
should be noted that the existing plugin interface was not designed with
parallel use in mind and this RFC does not claim to address that broader
limitation.

A further concern specific to floating-point parameters is
\textit{strtod} determinism on heterogeneous clusters.  On a cluster mixing
CPU architectures (e.g., x86\_64 and ARM64), the conversion of a decimal
string such as \texttt{"rate=3.55555555"} to a \texttt{double} may produce
different bit patterns if the platform \texttt{strtod} implementations are
not both correctly-rounded per IEEE~754.  Because \texttt{H5Z\_cd\_pack\_double}
is layout-deterministic only after the \texttt{double} value is in hand, a
per-rank difference in the string-to-double conversion step propagates into
\texttt{cd\_values} and causes a DCPL mismatch at the collective
\texttt{H5Dcreate} call.  Applications running on heterogeneous clusters
with floating-point filter parameters should obtain a single parsed DCPL on
one rank and broadcast it to all others (e.g., via serialisation through
\texttt{H5Pencode}/\texttt{H5Pdecode}) rather than having each rank
independently invoke \texttt{H5Pset\_filter\_params}.

\verysubsection{Diagnosing rank-inconsistent DCPLs}
A DCPL mismatch across MPI ranks at \texttt{H5Dcreate} time produces an
\texttt{H5E\_CANTCREATE} error in debug builds but can manifest as a hang
or silent data corruption in optimised builds depending on the MPI
implementation.  To aid diagnosis, the environment variable
\texttt{HDF5\_MPI\_CHECK\_DCPL} (proposed for a future release) would
cause the library to perform an \texttt{MPI\_Allreduce} hash of the
serialised DCPL before each collective \texttt{H5Dcreate} and emit a
warning if any rank's hash differs.  In the interim, applications can
replicate this check manually: serialise the DCPL with \texttt{H5Pencode}
on every rank, hash the result, \texttt{MPI\_Allreduce} the hashes, and
abort collectively before \texttt{H5Dcreate} if any hash differs.

\verysubsection{Failure mode and plugin loading order}

\noindent\textbf{Decision:} \texttt{H5Pset\_filter\_params} loads the plugin
eagerly at call time and invokes \texttt{set\_config} immediately.  Unlike
\texttt{H5Pset\_filter}, which stores raw \texttt{cd\_values} and defers
all plugin interaction, \texttt{H5Pset\_filter\_params} must run the filter's
\texttt{set\_config} callback to translate the parameter string into
\texttt{cd\_values}; this requires the plugin to be registered.  If the
plugin is unavailable, \texttt{H5Pset\_filter\_params} returns \texttt{H5E\_NOFILTER}
at the call site rather than deferring failure to \texttt{H5Dcreate} or
I/O time.  For Parallel HDF5, all ranks call \texttt{H5Pset\_filter\_params}
with identical arguments; a failure on any rank is handled locally before
any collective \texttt{H5Dcreate} is issued.  Applications that want to
probe filter availability before committing to a DCPL configuration should
call \texttt{H5Zfilter\_avail} first.

\subsection{VOL Connector Interaction}

Filters are applied to chunks by the native VOL connector during I/O.
\texttt{H5Pset\_filter\_params} resolves any parameter string into standard
\texttt{cd\_values} before modifying the DCPL. From the VOL connector's perspective,
the DCPL contains the same filter ID, flags, and \texttt{cd\_values} as if
\texttt{H5Pset\_filter} had been called directly.

Therefore, no changes to the VOL connector interface are required. Third-party
VOL connectors that intercept DCPLs will see standard filter configurations
regardless of whether they were set via \texttt{H5Pset\_filter\_params} or
\texttt{H5Pset\_filter}.

Parameter string resolution occurs at \texttt{H5Pset\_filter\_params}
call time, before any dataset is opened or created. It therefore has no
interaction with SWMR (which governs I/O access patterns on open datasets) or
with asynchronous VOL event sets (which are populated after the DCPL is fully
constructed).

\verysubsection{Note for pass-through and network VOL connectors}
VOL connectors that proxy dataset creation to a remote server (e.g., REST~VOL,
DAOS~VOL) receive a DCPL whose filter entries are already in \texttt{cd\_values}
form.  A remote server that natively understands key=value filter strings cannot
recover the original string from \texttt{cd\_values} without calling
\texttt{get\_config}.  This is an intentional trade-off: resolution at the
\texttt{H5Pset\_filter\_params} call site is the simplest design that keeps the
VOL interface unchanged.  Remote VOLs that require string-form parameters for
their own API can call \texttt{H5Pget\_filter\_params\_by\_idx} (which invokes
\texttt{get\_config}) to reconstruct a best-effort string from the stored
\texttt{cd\_values} before forwarding the request.

\textbf{Known limitation for thin-client architectures:}
The reconstruction path above requires the filter plugin to be installed on the
\emph{client} machine so that \texttt{get\_config} can be invoked.  For true
thin-client REST deployments where the client has no local copy of the plugin,
neither the original string nor a reconstructed equivalent is available after
\texttt{H5Pset\_filter\_params} returns.  This is a direct consequence of the
eager-parsing design and is the primary trade-off of resolving strings at
property-list configuration time rather than at dataset creation time on the
server.  Applications targeting thin-client REST VOLs that need to forward
filter parameters as strings should use \texttt{H5Pset\_filter} directly with
raw \texttt{cd\_values} constructed offline, or negotiate a separate
out-of-band configuration channel with the server.


%% ======================================================================
%%  Section 9 – CLI Tool Integration
%% ======================================================================
\section{CLI Tool Integration}
\label{sec:cli-tools}

\subsection{h5repack}

\texttt{h5repack} currently accepts filter specifications via the \texttt{-f} flag. The
existing syntax (\texttt{GZIP}, \texttt{SZIP}, \texttt{SHUF}, \texttt{FLET}, \texttt{NBIT},
\texttt{SOFF}, and \texttt{UD=id,flags,n,v1,...}) continues to work unchanged.

A new form is added that accepts a filter ID and parameter string:

\begin{Verbatim}[fontsize=\scriptsize]
    h5repack -f 'UD=1,0,level=6' input.h5 output.h5
    h5repack -f 'UD=32013,0,mode=rate,rate=3.5' input.h5 output.h5
    h5repack -f '/group/dataset:UD=32013,0,mode=rate,rate=3.5' input.h5 output.h5
\end{Verbatim}

The existing \texttt{UD=id,flags,...} syntax is extended to accept a parameter
string in place of the raw \texttt{cd\_values} sequence when the filter is a v3
plugin that implements \texttt{set\_config}. The \texttt{h5repack} parser detects
whether the values following the flags field are a key=value string or a raw
integer sequence and dispatches accordingly---calling \texttt{H5Pset\_filter\_params}
with \texttt{H5Z\_PARAMS\_STR} for the string form and \texttt{H5Z\_PARAMS\_RAW}
for the raw integer form.
The old built-in aliases (\texttt{GZIP}, \texttt{SHUF}, \texttt{FLET}, \texttt{SOFF})
continue to work unchanged.

\verysubsection{Duplicate filter entries}
HDF5 allows the same filter ID to appear more than once in a pipeline.
When \texttt{h5repack} applies a \texttt{-f} specification that matches a
filter already present in the source pipeline, it appends a new entry
rather than replacing the existing one---this is the same behaviour as the
existing \texttt{UD=} raw-integer form.  Pipeline ordering is fully
preserved: entries are written to the output DCPL in the order they appear
in the source pipeline, followed by any new entries added via \texttt{-f}.
No new behaviour is introduced for this edge case.

\subsection{h5dump}

The \texttt{PARAMS\_STRING} field is displayed only when \texttt{h5dump} is invoked
with \texttt{-p} (the existing flag that enables display of filter and property
details). Output without \texttt{-p} is unchanged. Because \texttt{PARAMS\_STRING} is
gated behind \texttt{-p}, and because format changes of this kind have been
accepted in prior major releases, no additional command-line flag is required.

\verysubsection{Normative output format (with \texttt{-p})}

When \texttt{-p} is active, for each filter in the pipeline that
implements \texttt{get\_config}, a \texttt{PARAMS\_STRING} line is appended immediately
after the existing filter line, indented to the same level as other filter
sub-fields:

\begin{verbatim}
    FILTERS {
      COMPRESSION DEFLATE { LEVEL 6 }
      PARAMS_STRING "level=6"
    }
\end{verbatim}

The value is the literal output of \texttt{get\_config} enclosed in double quotes.
If the string returned by \texttt{get\_config} contains a double-quote character,
it is escaped as \texttt{\textbackslash\textquotedbl} in the h5dump output. For filters that do not implement \texttt{get\_config},
\texttt{H5Pget\_filter\_params\_by\_idx} provides a fallback string that formats the
raw \texttt{cd\_values} as a colon-separated list (see \SectionRef{sec:introspection}),
so \texttt{PARAMS\_STRING} can be displayed for all filters regardless of whether
they implement \texttt{get\_config}.

The format change is confined to the \texttt{FILTERS} block; all other
\texttt{h5dump} output is unaffected.


%% ======================================================================
%%  Section 10 – Language Bindings
%% ======================================================================
\section{Language Bindings}
\label{sec:bindings}

\subsection{Fortran}
\label{sec:fortran-bindings}

Fortran's ISO\_C\_BINDING interop cannot represent a C union, so
\texttt{H5Z\_params\_t} cannot be expressed as a \texttt{BIND(C)} derived
type.  Two thin C helper functions are therefore added to
\texttt{H5Zf.c} --- one per calling mode --- each of which constructs the
appropriate \texttt{H5Z\_params\_t} value and calls
\texttt{H5Pset\_filter\_params}:

\begin{minted}{c}
/* In H5Zf.c -- C-side entry points for Fortran BIND(C) */

herr_t H5Pset_filter_params_str_c(hid_t plist, H5Z_filter_t id,
                                   unsigned flags, const char *params)
{
    H5Z_params_t p = H5Z_PARAMS_STR(params);
    return H5Pset_filter_params(plist, id, flags, &p);
}

herr_t H5Pset_filter_params_raw_c(hid_t plist, H5Z_filter_t id,
                                   unsigned flags, size_t cd_nelmts,
                                   const unsigned *cd_values)
{
    H5Z_params_t p = H5Z_PARAMS_RAW(cd_nelmts, cd_values);
    return H5Pset_filter_params(plist, id, flags, &p);
}
\end{minted}

The Fortran binding then exposes a generic interface
\texttt{h5pset\_filter\_params\_f} with two specific procedures that bind
directly to these C entry points via \texttt{BIND(C)}.  Fortran's generic
resolution selects the correct specific at the call site based on the type
of the actual argument (\texttt{CHARACTER} vs.\ \texttt{INTEGER} array).

\begin{minted}{fortran}
! In H5Pff.F90 -- generic interface
INTERFACE h5pset_filter_params_f
  MODULE PROCEDURE h5pset_filter_params_str_f   ! string form
  MODULE PROCEDURE h5pset_filter_params_raw_f   ! cd_values form
END INTERFACE

! String form -- binds directly to H5Pset_filter_params_str_c
SUBROUTINE h5pset_filter_params_str_f( &
    prp_id, filter_id, flags, params, hdferr) &
    BIND(C, NAME='H5Pset_filter_params_str_c')
  USE ISO_C_BINDING
  INTEGER(HID_T),        INTENT(IN)  :: prp_id
  INTEGER,               INTENT(IN)  :: filter_id
  INTEGER,               INTENT(IN)  :: flags
  CHARACTER(KIND=C_CHAR),INTENT(IN)  :: params(*)
  INTEGER,               INTENT(OUT) :: hdferr

! Raw cd_values form -- binds directly to H5Pset_filter_params_raw_c
SUBROUTINE h5pset_filter_params_raw_f( &
    prp_id, filter_id, flags, cd_nelmts, cd_values, hdferr) &
    BIND(C, NAME='H5Pset_filter_params_raw_c')
  USE ISO_C_BINDING
  INTEGER(HID_T),        INTENT(IN)  :: prp_id
  INTEGER,               INTENT(IN)  :: filter_id
  INTEGER,               INTENT(IN)  :: flags
  INTEGER(SIZE_T),       INTENT(IN)  :: cd_nelmts
  INTEGER,               DIMENSION(*), INTENT(IN) :: cd_values
  INTEGER,               INTENT(OUT) :: hdferr
\end{minted}

\verysubsection{Fortran string passing convention}
Fortran \texttt{CHARACTER(LEN=*)} strings are not NUL-terminated.  The
C helper \texttt{H5Pset\_filter\_params\_str\_c} receives
\texttt{CHARACTER(KIND=C\_CHAR)} and therefore expects a NUL-terminated
string.  The caller is responsible for appending \texttt{C\_NULL\_CHAR}
when constructing the actual argument, following the same convention used
throughout the existing HDF5 Fortran binding:

\begin{minted}{fortran}
CALL h5pset_filter_params_f(plist, H5Z_FILTER_DEFLATE_F, &
                             H5Z_FLAG_MANDATORY_F,        &
                             "level=6"//C_NULL_CHAR, hdferr)
\end{minted}

The following additional subroutines are added to the Fortran binding.
Each binds directly to a corresponding thin C helper in \texttt{H5Zf.c}
that handles any \texttt{CHARACTER} $\leftrightarrow$ \texttt{char~*}
translation, following the same pattern as above.

\begin{minted}{fortran}
! Pipeline introspection -- binds to H5Pget_filter_params_by_idx_c
SUBROUTINE h5pget_filter_params_by_idx_f( &
    prp_id, filter_idx, params_buf, params_len, hdferr) &
    BIND(C, NAME='H5Pget_filter_params_by_idx_c')
  USE ISO_C_BINDING
  INTEGER(HID_T),        INTENT(IN)    :: prp_id
  INTEGER,               INTENT(IN)    :: filter_idx
  CHARACTER(KIND=C_CHAR),INTENT(OUT)   :: params_buf(*)
  INTEGER(SIZE_T),       INTENT(INOUT) :: params_len
  INTEGER,               INTENT(OUT)   :: hdferr

! Parameter lookup helper -- binds to H5Zconfig_get_param_c
FUNCTION h5zconfig_get_param_f( &
    params, key, value_buf, buf_size) RESULT(found) &
    BIND(C, NAME='H5Zconfig_get_param_c')
  USE ISO_C_BINDING
  CHARACTER(KIND=C_CHAR),INTENT(IN)    :: params(*)
  CHARACTER(KIND=C_CHAR),INTENT(IN)    :: key(*)
  CHARACTER(KIND=C_CHAR),INTENT(OUT)   :: value_buf(*)
  INTEGER(SIZE_T),       INTENT(INOUT) :: buf_size
  INTEGER(C_INT)                       :: found   ! htri_t: >0 found, 0 absent, <0 error
\end{minted}

\verysubsection{Fortran usage example}

\begin{minted}{fortran}
USE HDF5
USE ISO_C_BINDING
IMPLICIT NONE
INTEGER(HID_T)        :: plist
INTEGER               :: hdferr
INTEGER, DIMENSION(1) :: cd_vals = [6]

CALL h5pcreate_f(H5P_DATASET_CREATE_F, plist, hdferr)

! String form (generic resolves to h5pset_filter_params_str_f)
CALL h5pset_filter_params_f(plist, H5Z_FILTER_DEFLATE_F, &
                             H5Z_FLAG_MANDATORY_F,        &
                             "level=6"//C_NULL_CHAR, hdferr)

! Raw cd_values form (generic resolves to h5pset_filter_params_raw_f)
CALL h5pset_filter_params_f(plist, H5Z_FILTER_DEFLATE_F, &
                             H5Z_FLAG_MANDATORY_F,        &
                             1_SIZE_T, cd_vals, hdferr)
\end{minted}


\subsection{Java}
\label{sec:java-bindings}

Java does not support C unions, and method overloading is the idiomatic
Java mechanism for variant inputs.  Two overloaded static methods are
added to the existing \texttt{HDF5Library} class:

\begin{minted}{java}
// String form
public static native int H5Pset_filter_params(
    long   plist_id,
    int    filter_id,
    int    flags,
    String params       // null means "no parameters"
) throws HDF5LibraryException, NullPointerException;

// Raw cd_values form (equivalent to existing H5Pset_filter)
public static native int H5Pset_filter_params(
    long   plist_id,
    int    filter_id,
    int    flags,
    int[]  cd_values    // null or zero-length means "no parameters"
) throws HDF5LibraryException;
\end{minted}

Each JNI implementation (in \texttt{H5P.c} on the native side) constructs
the appropriate \texttt{H5Z\_params\_t} value using \texttt{H5Z\_PARAMS\_STR}
or \texttt{H5Z\_PARAMS\_RAW} and calls \texttt{H5Pset\_filter\_params}.
The \texttt{String} overload converts the Java string to a NUL-terminated
\texttt{const char~*} via \texttt{GetStringUTFChars} before constructing
the \texttt{H5Z\_params\_t}.

The following additional method is added for pipeline introspection:

\begin{minted}{java}
// Wraps H5Pget_filter_params_by_idx
public static native String H5Pget_filter_params_by_idx(
    long plist_id,
    int  filter_idx
) throws HDF5LibraryException;
\end{minted}

\verysubsection{Java usage example}

\begin{minted}{java}
import hdf.hdf5lib.H5;
import hdf.hdf5lib.HDF5Constants;

long plist = H5.H5Pcreate(HDF5Constants.H5P_DATASET_CREATE);

// String form
H5.H5Pset_filter_params(plist, HDF5Constants.H5Z_FILTER_DEFLATE,
                         HDF5Constants.H5Z_FLAG_MANDATORY, "level=6");

// Raw cd_values form
H5.H5Pset_filter_params(plist, HDF5Constants.H5Z_FILTER_DEFLATE,
                         HDF5Constants.H5Z_FLAG_MANDATORY,
                         new int[]{ 6 });
\end{minted}


\subsection{C++}
\label{sec:cpp-bindings}

The HDF5 C++ wrapper library (\texttt{H5Cpp}) wraps C functions as methods
on \texttt{DSetCreatPropList} and related classes.  C++ does not support C99
compound literals, so \texttt{H5Z\_PARAMS\_STR} and \texttt{H5Z\_PARAMS\_RAW}
expand to C++-compatible brace-initialisers under \texttt{\#ifdef \_\_cplusplus}
(\SectionRef{sec:set-filter-params}).  The named-variable form is preferred
for clarity:

\begin{minted}{cpp}
// Preferred named-variable form in C++
H5Z_params_t p;
p.type  = H5Z_PARAMS_STRING;
p.u.str = "level=6";
H5Pset_filter_params(plist.getId(), H5Z_FILTER_DEFLATE,
                     H5Z_FLAG_MANDATORY, &p);
\end{minted}

A thin C++ wrapper method is added to \texttt{DSetCreatPropList}:

\begin{minted}{cpp}
// In DSetCreatPropList.h
class DSetCreatPropList : public PropList {
public:
    // String form
    void setFilterParams(H5Z_filter_t filter_id, unsigned flags,
                         const std::string &params);

    // Raw cd_values form (equivalent to existing setFilter)
    void setFilterParams(H5Z_filter_t filter_id, unsigned flags,
                         size_t cd_nelmts, const unsigned cd_values[]);
};
\end{minted}

Each overload constructs an \texttt{H5Z\_params\_t} on the stack and calls
\texttt{H5Pset\_filter\_params}.  No new C entry point is required.

\begin{minted}{cpp}
// C++ usage example
DSetCreatPropList dcpl;

// String form
dcpl.setFilterParams(H5Z_FILTER_DEFLATE, H5Z_FLAG_MANDATORY, "level=6");

// Raw cd_values form
unsigned cd[] = { 6 };
dcpl.setFilterParams(H5Z_FILTER_DEFLATE, H5Z_FLAG_MANDATORY, 1, cd);
\end{minted}

\end{document}
