\documentclass[../RFC-HDFG-2026-001.tex]{subfiles}

\begin{document}

%% ======================================================================
%%  Section 12 – Resolved Design Decisions
%% ======================================================================
\section{Resolved Design Decisions}
\label{sec:resolved}

The following items were raised during review and have been resolved.

\verysubsection{R1}
Introspection API naming and signature. The introspection API went through
several iterations: (1) a single \texttt{ssize\_t}-returning function with two
output buffers and ambiguous return semantics; (2) two \texttt{ssize\_t}-returning
functions including a name-by-index variant; (3) a single merged \texttt{herr\_t}
function with nine parameters. The final design uses one focused
\texttt{herr\_t}-returning function---\texttt{H5Pget\_filter\_params\_by\_idx}
(5~params)---with an explicit \texttt{size\_t*} output for string length.
A name-by-index variant (\texttt{H5Pget\_filter\_name\_by\_idx}) was considered
but dropped: for external filters it would return only the numeric ID (since
v2 plugins and unregistered filters have no name in the registry), making the
function unreliable across the general filter population.
The \texttt{filter\_idx} parameter uses \texttt{unsigned} to match \texttt{H5Pget\_filter2}. See
\SectionRef{sec:introspection}.

\verysubsection{R2}
Memory bounds for \texttt{cd\_values}. The \texttt{set\_config} callback supports a
two-pass size query pattern: the library first calls with \texttt{cd\_values=NULL} to
get the required count, allocates accordingly, and calls again. See
\SectionRef{sec:callback-specs}.

\verysubsection{R3}
Raw fallback syntax in the string API. The library does NOT implement a
special \texttt{cd\_values=...} pass-through syntax within
\texttt{H5Pset\_filter2}. Users who need to pass raw \texttt{cd\_values} to a v2
plugin should use the existing \texttt{H5Pset\_filter} API. This keeps the string
API clean.

\verysubsection{R4}
Pipeline-in-a-string as primary API. The single-string pipeline
variant (\SectionRef{sec:pipeline-string}) is NOT the primary API. The primary
API is \texttt{H5Pset\_filter2} (one filter per call, appending), consistent
with established HDF5 conventions.

\verysubsection{R5}
DCPL string storage. The DCPL does NOT retain the original parameter
string. After resolution, only the standard \texttt{cd\_values}, filter ID, and flags
are stored. Introspection functions rely on \texttt{get\_config} to reconstruct a
human-readable string from \texttt{cd\_values}.

\verysubsection{R6}
VOL connector interaction. The string API resolves to \texttt{cd\_values}
before the VOL connector sees the DCPL. No changes to the VOL interface are
required. See \SectionRef{sec:thread-safety}.

\verysubsection{R7}
Name-based plugin discovery. Plugin loading uses the existing integer-ID
path (\texttt{H5PL}). The \texttt{H5PL\_filter\_key\_t} discriminated-union extension and
\texttt{H5Z\_check\_plugin\_load()} callback described in an earlier draft are NOT
part of this release---they are deferred pending adoption of a name-based
primary API. Plugin loading is always by integer ID; the name registry
(\SectionRef{sec:name-registry}) does not trigger plugin loading.  A plugin
must already be registered (either as a built-in or via a prior
\texttt{H5Zregister} / \texttt{H5PL} load) before its name appears in the
registry.  The registry is used by \texttt{h5repack} to resolve canonical
filter names to integer IDs before calling \texttt{H5Pset\_filter2}; this
lookup performs no dynamic library loading.

\verysubsection{R8}
Version numbering for \texttt{H5Z\_class3\_t}. The new struct carries
\texttt{version = 2}. The dispatch logic in \texttt{H5Zregister()} is extended to a
three-way branch (two-way under \texttt{H5\_NO\_DEPRECATED\_SYMBOLS}). The
\texttt{H5Z\_class\_t\_vers} compatibility macro in \texttt{H5version.h} retains its default
of~2 (mapping to \texttt{H5Z\_class2\_t}); it is NOT bumped to~3 to avoid silent
struct-size mismatches. See \SectionRef{sec:version-numbering}.

\verysubsection{R9}
Built-in filter validation. \texttt{set\_config} callbacks for built-in
filters with non-trivial validation (SZIP, Scale-Offset) delegate to existing
internal validation code rather than reimplementing it. See
\SectionRef{sec:builtin-filters}.

\verysubsection{R10}
Error code for unsupported string config. The error
\texttt{H5E\_UNSUPPORTED} (not a new code) is used when a filter does not implement
\texttt{set\_config}, rather than a novel \texttt{H5E\_CANTENCODE} code. See
\SectionRef{sec:set-filter2}.

\verysubsection{R11}
On-demand plugin loading path. Plugin loading is always by integer ID via
the existing \texttt{H5PL} infrastructure. The name registry
(\SectionRef{sec:name-registry}) is populated during plugin registration and
is used by \texttt{h5repack} to resolve canonical names to filter IDs; it has
no role in triggering plugin loading.
\texttt{H5Zfilter\_avail} (by integer ID) is the correct API for pre-validating
filter availability before committing to a DCPL configuration.

\verysubsection{R12}
Introspection fallback for filters without \texttt{get\_config}.
\texttt{H5Pget\_filter\_params\_by\_idx} formats the raw \texttt{cd\_values} as a fallback
string using colon separators (e.g., \texttt{"cd\_values=1:0:0:1074921472"}) so that
introspection always yields something useful. The colon separator avoids
ambiguity with the comma-separated grammar and keeps the fallback string valid
under the grammar as a single key=value pair. See \SectionRef{sec:introspection}.

\verysubsection{R13}
Name registry write path. There is no public API for registering name aliases
(\texttt{H5Zregister\_name} was considered and rejected). Names enter the registry
exclusively through built-in registration at library initialization and v3
plugin self-registration via \texttt{H5Z\_class3\_t.name}. Introspection returns the
first-registered name for a given filter ID. See \SectionRef{sec:name-registry}.

\verysubsection{R14}
No numeric-string disambiguation needed. \texttt{H5Pset\_filter2} takes an
explicit \texttt{H5Z\_filter\_t} integer ID rather than a string name, eliminating
the ambiguity between a numeric filter-ID string and a registered canonical
name that was present in the earlier name-based API design. See
\SectionRef{sec:set-filter2}.

\verysubsection{R15}
\texttt{H5Z\_CLASS\_T\_VERS} and \texttt{H5Z\_class\_t\_vers} must not change. The
existing constant \texttt{H5Z\_CLASS\_T\_VERS} (= 1) is left unchanged. A new constant
\texttt{H5Z\_CLASS3\_T\_VERS} (= 2) is introduced for v3 structs. The
\texttt{H5Z\_class\_t\_vers} compatibility macro in \texttt{H5version.h} retains its default
value of~2 (mapping \texttt{H5Z\_class\_t} $\rightarrow$ \texttt{H5Z\_class2\_t}) and is NOT
bumped to~3. See \SectionRef{sec:version-numbering}.

\verysubsection{R16}
Developer helper for plugin authors. The library provides
\texttt{H5Zconfig\_get\_param} in \texttt{H5Zdevelop.h}, returning \texttt{htri\_t} with an
explicit \texttt{size\_t *val\_len} output. See \SectionRef{sec:config-get-param}.

\verysubsection{R17}
Filter availability check. A by-name variant of \texttt{H5Zfilter\_avail} is NOT
provided. The existing \texttt{H5Zfilter\_avail(id)} function (by integer ID) is the
correct API for checking filter availability before committing to a DCPL
configuration.

\verysubsection{R18}
Duplicate filter IDs. A filter ID never appears more than once in a
pipeline; it is the caller's responsibility not to add the same filter ID
twice. \texttt{H5Pset\_filter2} appends to the pipeline, matching
\texttt{H5Pset\_filter}. See \SectionRef{sec:set-filter2}.

\verysubsection{R19}
Grammar forward compatibility. Future extensions to the parameter
string grammar are backward-compatible: any string valid under the current
grammar remains valid with the same semantics under future grammars. See
\SectionRef{sec:risks}.

\verysubsection{R20}
Parameterless filter rejection. \texttt{set\_config} callbacks for filters
that take no parameters (shuffle, fletcher32, nbit) should reject
non-\texttt{NULL}, non-empty \texttt{params} strings with \texttt{H5E\_BADVALUE}. See
\SectionRef{sec:set-filter2}.

\verysubsection{R21}
\texttt{get\_config} round-tripping. The output of \texttt{get\_config} conforms
to the same grammar as \texttt{set\_config} input (using \texttt{key=value}
pairs and quoted values where appropriate) so that \texttt{get\_config} output
is valid \texttt{set\_config} input. See \SectionRef{sec:callback-specs}.

\verysubsection{R22}
Introspection return convention. \texttt{H5Pget\_filter\_params\_by\_idx} returns
\texttt{herr\_t} with an explicit \texttt{size\_t*} output parameter for reporting the
string length, following the newer HDF5 API convention. See
\SectionRef{sec:introspection}.

\verysubsection{R24}
Two-pass \texttt{cd\_nelmts} identity contract. The \texttt{set\_config} callback
returns the same \texttt{cd\_nelmts} on both passes of the size-query protocol. If
the second-pass value differs from the first (in either direction), the
library returns an error and does not write to the \texttt{cd\_values} buffer
with the inconsistent count. A smaller second-pass value would silently
truncate the populated array; a larger value risks a buffer overflow. Both
cases indicate a buggy callback. The RFC previously specified only the
overflow guard (second-pass exceeds allocated capacity). This resolved item
adds the symmetric lower-bound check and makes identical counts the explicit
contract. See \SectionRef{sec:callback-specs}.

\verysubsection{R25}
Empty keys forbidden; bare keys allowed; explicit empty value rejected. The
grammar forbids empty keys (\textit{key} requires one or more characters) and
this constraint is stated explicitly in the Rules section. Bare keys (tokens
without an equals sign, e.g., \texttt{"fast\_mode"}) ARE permitted and are
treated as boolean flags with an implicit empty value;
\texttt{H5Zconfig\_get\_param} returns positive with \texttt{*buf\_size = 0}
for a bare key. A token with an equals sign but no value (\texttt{"key="})
is rejected with \texttt{H5E\_BADVALUE}. The string \texttt{"=value"}
is also rejected with \texttt{H5E\_BADVALUE}.
See \SectionRef{sec:param-rules}.

\verysubsection{R26}
\texttt{h5repack} \texttt{UD=} disambiguation. When parsing the new
\textit{filtername}\texttt{=}\textit{params} syntax, a filter name of exactly \texttt{UD}
(case-insensitive) is ambiguous because the legacy \texttt{UD=id,flags,n,v1,...}
syntax uses the same prefix. A precedence rule is added: if the text
following \texttt{UD=} begins with a decimal digit, the specification is parsed
as the legacy syntax; otherwise it is treated as new-syntax where \texttt{UD} is a
registered canonical filter name. See \SectionRef{sec:cli-tools}.

\verysubsection{R27}
\texttt{h5dump} \texttt{PARAMS\_STRING} format change. The addition of
\texttt{PARAMS\_STRING} to \texttt{h5dump} output constitutes a format change that may
break downstream parsers. The field is gated behind a command-line option
(\texttt{--filter-params}) so that the default output is unchanged. The output
format when the option is active is specified normatively in the CLI Tools
section. See \SectionRef{sec:cli-tools}.

\verysubsection{R23}
Flags-override warning contract. Rather than silently overriding the
caller's \texttt{flags} value, a \texttt{set\_config} callback that modifies \texttt{flags}
pushes a non-fatal warning via \texttt{H5Epush} with severity \texttt{H5E\_WARN} and class
\texttt{H5E\_ARGS} before returning. Callbacks that accept the caller's flags
without modification do not push this warning. This allows callers who want
to detect a flag override to inspect the error stack after
\texttt{H5Pset\_filter2} returns, while callers that do not check the stack
are entirely unaffected---preserving the existing SZIP behavior of
unconditionally forcing \texttt{H5Z\_FLAG\_OPTIONAL}. Document this contract in
\texttt{H5Zdevelop.h} as part of the \texttt{H5Z\_set\_config\_func\_t}
contract. See \SectionRef{sec:callback-specs} and \SectionRef{sec:builtin-filters}.

\verysubsection{R32}
API chosen: \texttt{H5Pset\_filter2} (ID-based with string params). The
working group decided to introduce a versioned variant \texttt{H5Pset\_filter2} rather
than a name-based \texttt{H5Pset\_filter\_by\_name}. Taking the filter ID from the
caller eliminates name-to-ID resolution from the public library API and
name-based plugin discovery, along with the associated registry write-path
complexity.  Name-to-ID resolution is still performed by \texttt{h5repack}
internally, using the library's name registry, before calling
\texttt{H5Pset\_filter2}; this is necessary because external filter plugins
can only be addressed by a name that was registered when the plugin was
loaded---there is no other way for \texttt{h5repack} to obtain the integer ID
for an external filter at runtime. The cost is the full HDF5 versioning
machinery (\texttt{H5version.h} compatibility macros, Fortran/Java bindings,
deprecation handling) for the \texttt{H5Pset\_filter2} name, which the group judged
acceptable given the long-term usability benefit. This decision also simplifies
the new additions to the public \texttt{H5Z} API surface: the only new public
functions are \texttt{H5Pset\_filter2}, \texttt{H5Pget\_filter\_params\_by\_idx},
and the developer-facing \texttt{H5Zconfig\_get\_param}. See \SectionRef{sec:set-filter2}.

\verysubsection{R28}
Normative string and parameter limits. The maximum parameter string length
(4096~bytes, \texttt{H5Z\_CONFIG\_STRING\_MAX}) and the maximum number of
parameters per string (64, \texttt{H5Z\_CONFIG\_MAX\_PARAMS}) are hard limits
enforced by the library, not recommendations. Both
constants are public. See \SectionRef{sec:param-rules}.

\verysubsection{R29}
Locale independence for both input and output. The library's parameter
string parser, \texttt{set\_config} callbacks, and \texttt{get\_config} callbacks
all operate in C-locale semantics. Process-wide locale settings do not
affect parsing or formatting. This ensures round-trip correctness in
environments with non-ASCII-compatible locale settings. See
\SectionRef{sec:param-rules} and \SectionRef{sec:callback-specs}.

\verysubsection{R30}
Canonical \texttt{set\_config} implementation template. A normative code
template is provided in the architecture section to guide plugin authors in
correctly implementing the two-pass \texttt{set\_config} protocol. All built-in
filter \texttt{set\_config} callbacks follow this structure. See
\SectionRef{sec:set-config-template}.

\verysubsection{R31}
\texttt{filter\_title} field in \texttt{H5Z\_class3\_t}. A new optional
\texttt{filter\_title} field provides a short human-readable label for the
filter. When non-\texttt{NULL}, the library packs it into \texttt{cd\_values}
as a reserved key so it persists in the file. \texttt{H5Pget\_filter2} extracts
\texttt{filter\_title} from \texttt{cd\_values} when present, falling back to
\texttt{name} otherwise. This mitigates the semantic change in the \texttt{name}
field and allows tools to display a human-readable label even when the plugin
is not installed. See
\SectionRef{sec:class3-struct} and \SectionRef{sec:name-semantic-change}.

\verysubsection{R35}
On-disk format extended via \texttt{H5O\_PLINE\_VERSION\_3}
(\SectionRef{sec:pline-v3}).  The decision to include the format change in
this RFC (rather than deferring) reflects two realities: (1)~HDF5 does not
guarantee forward compatibility, so the standard mechanism for user-controlled
format selection is \texttt{H5Pset\_libver\_bounds}, which already gates all
other format extensions; (2)~stopping at string-in/uint32-on-disk leaves
parameters opaque to tools that cannot load the plugin, which contradicts the
self-describing goal of HDF5.  Files written with a libver upper bound below
\texttt{H5F\_LIBVER\_V300} are byte-identical to \texttt{H5Pset\_filter} output.
The \texttt{H5Z\_class4\_t} schema extension (\SectionRef{sec:class4}) remains
informative and is deferred to a follow-on RFC.
See \SectionRef{sec:introspection}.

\verysubsection{R36}
\texttt{H5Pget\_filter2} semantics for v3 plugins.  For v3 plugins,
\texttt{H5Pget\_filter2} extracts \texttt{filter\_title} from \texttt{cd\_values}
when the key is present, falling back to \texttt{name}.  This preserves
backward-compatible human-readable output provided plugin authors populate
\texttt{filter\_title} on migration, and works even when the plugin is not
loaded since the title is persisted in the file.
\texttt{H5Pget\_filter\_params\_by\_idx} (parameter string via
\texttt{get\_config}) is the new introspection function.
See \SectionRef{sec:introspection}.

\verysubsection{R37}
\texttt{H5Pset\_filter2} loads the plugin eagerly at call time.  Unlike
\texttt{H5Pset\_filter}, \texttt{H5Pset\_filter2} must invoke the filter's
\texttt{set\_config} callback immediately to translate the parameter string
into \texttt{cd\_values}; this requires the plugin to be registered.  A
missing plugin returns \texttt{H5E\_NOFILTER} at the call site, not at
\texttt{H5Dcreate} time, giving earlier and more diagnosable errors.
For Parallel HDF5, all ranks call \texttt{H5Pset\_filter2} before any
collective \texttt{H5Dcreate}.  See \SectionRef{sec:set-filter2}.

\verysubsection{R38}
Raw-string callback interface retained.  \texttt{set\_config} receives
\texttt{const char *params} and uses \texttt{H5Zconfig\_get\_param} for
key lookup.  Pre-tokenization was rejected: parsing overhead is
microseconds vs.\ millisecond dataset creation; switching would change the
\texttt{H5Z\_set\_config\_func\_t} ABI and burden plugin authors.  Token
caching can be added transparently inside the library if ever needed.
See \SectionRef{sec:callback-specs}.

\verysubsection{R34}
\texttt{H5Z\_class3\_t} struct layout --- embedding rejected.
Three approaches were considered: (A)~embed \texttt{H5Z\_class2\_t} as the
first member of \texttt{H5Z\_class3\_t}; (B)~add an optional
\texttt{H5PLget\_plugin\_version()} plugin symbol; (C)~retain a flat
independent struct and resolve the strict-aliasing issue in
\texttt{H5Zregister()} and \texttt{H5PL\_\_open} via \texttt{memcpy}
through \texttt{char~*}.  The working group selected option~(C) at the
April~23, 2026 design review.  Struct embedding was rejected because it is
inconsistent with HDF5 versioning conventions and would make
\texttt{H5Z\_class3\_t} depend on a deprecated type under
\texttt{H5\_NO\_DEPRECATED\_SYMBOLS}.  The version-query callback was
rejected because it imposes a new per-plugin symbol and does not cover the
\texttt{H5Zregister()} direct-registration path.
See \SectionRef{sec:version-numbering}.

\verysubsection{R33}
Duplicate key handling. Duplicate keys in a parameter string are rejected
with \texttt{H5E\_BADVALUE}.  A survey of related technologies shows two
camps: last-wins (CMake \texttt{set()}~\cite{cmakeset}, most JSON
parsers~\cite{rfc8259}) and strict rejection (TOML~\cite{toml10},
HTTP RFC~9110~\cite{rfc9110}, Python \texttt{configparser}
\texttt{strict=True}~\cite{pythonconfigparser}).  For a filter configuration
string, a duplicate key is almost certainly a user error; silent last-wins
would mask it.  Detection overhead is bounded by
\texttt{H5Z\_CONFIG\_MAX\_PARAMS}\,$=\,64$ comparisons per parse.
See \SectionRef{sec:param-rules}.

\end{document}
